---
title: "Public Health and Economic Impacts of Severe Weather Events in the US"
author: "Brian High"
date: "01/25/2015"
output: html_document
---

## Synopsis

To determine the primary impacts of weather events on US health and economic
wellbeing, we will analyze data from the 
[NOAA National Climatic Data Center](http://www.ncdc.noaa.gov/stormevents/). 

The data includes counts by 
storm event type for injuries, fatalities, property damage and crop 
damage. We will compare totals of these counts for the entire time period 
to determine the top five weather event impacts. Then we will group these 
totals by month to plot the seasonal occurrence of these events. By seeing 
the most significant weather events and their impacts over the course of the 
year, decisions can be made to allocate resources to better manage these 
impacts.

Across the United States, the types of events which are the most harmful 
with respect to population health are tornados, heat, thunderstorms, 
lightning, and flooding. Floods, hurricanes, typhoons, tornadoes, storm 
surge, and hail have the greatest economic consequences.

## Data Processing

### Load and preprocess the data

Download the compressed CSV file and load it into a data.table. Remove extra 
columns, format the date columns, add a Month column, and convert property 
and crop damage columns to USD.

```{r, echo = TRUE, cache=TRUE}
# Download the data file (if missing)
url <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
zipfile <- "StormData.csv.bz2"
if (! file.exists(zipfile)) { 
    download.file(url, zipfile , "auto") 
}

# Load into a data table
library(data.table)
stormdata <- as.data.table(read.csv(bzfile(zipfile), header = TRUE, 
                      stringsAsFactors = FALSE))

# Remove extra columns
stormdata[, c("STATE__","BGN_TIME","TIME_ZONE","COUNTY","COUNTYNAME",
              "BGN_RANGE","BGN_AZI","BGN_LOCATI","END_TIME","COUNTY_END",
              "COUNTYENDN","END_RANGE","END_AZI","END_LOCATI","LENGTH",
              "WIDTH","F","MAG","WFO","STATEOFFIC","ZONENAMES","LATITUDE",
              "LONGITUDE","LATITUDE_E","LONGITUDE_","REMARKS","REFNUM") := NULL]

# Prepare date columns
suppressMessages(library(lubridate))
stormdata[, c("BGN_DATE", "END_DATE") 
                    := list(mdy_hms(BGN_DATE), mdy_hms(END_DATE))]
stormdata[, "Month" := as.factor(month(BGN_DATE))]

# Convert property damage amounts to dollars
stormdata[, PROPDMGEXP := gsub("^[^HKMBhkmb]*$", 0, PROPDMGEXP)]
stormdata[, PROPDMGEXP := gsub("^[Hh]{1}$", 2, PROPDMGEXP)]
stormdata[, PROPDMGEXP := gsub("^[Kk]{1}$", 3, PROPDMGEXP)]
stormdata[, PROPDMGEXP := gsub("^[Mm]{1}$", 6, PROPDMGEXP)]
stormdata[, PROPDMGEXP := gsub("^[Bb]{1}$", 9, PROPDMGEXP)]
stormdata[, PROPDMG := as.numeric(PROPDMG)*10^as.numeric(PROPDMGEXP)]

# Convert crop damage amounts to dollars
stormdata[, CROPDMGEXP := gsub("^[^HKMBhkmb]*$", 0, CROPDMGEXP)]
stormdata[, CROPDMGEXP := gsub("^[Hh]{1}$", 2, CROPDMGEXP)]
stormdata[, CROPDMGEXP := gsub("^[Kk]{1}$", 3, CROPDMGEXP)]
stormdata[, CROPDMGEXP := gsub("^[Mm]{1}$", 6, CROPDMGEXP)]
stormdata[, CROPDMGEXP := gsub("^[Bb]{1}$", 9, CROPDMGEXP)]
stormdata[, CROPDMG := as.numeric(CROPDMG)*10^as.numeric(CROPDMGEXP)]
```

### Public Health Effects

Filter fatalities and injuries for 50 US states and group by event type.

```{r, echo = TRUE, cache=TRUE}
library(datasets)
health <- stormdata[STATE %in% state.abb, .(FATALITIES, INJURIES, Month), 
                    by=EVTYPE]
```

#### Top 5 Weather Event Types for Injuries

Determine the top five weather event types for injuries with a sum of all 
injuries, grouped by event type, for the entire time period.

```{r, echo = TRUE, cache=TRUE}
injuries <- health[,.(Injuries=sum(INJURIES)), EVTYPE]
setkey(injuries, Injuries)
topfiveinjuries <- tail(injuries, 5)
topfiveinjuries[order(-Injuries)]
```

From the list of top five weather event types for injuries, calculate injury 
sums by event type and month.

```{r, echo = TRUE, cache=TRUE}
injuriesbymonth <- health[EVTYPE %in% topfiveinjuries$EVTYPE,
                        .(Injuries=sum(INJURIES)), .(EVTYPE, Month)]
injuriesbymonth[, "EventType" := factor(EVTYPE, levels=rev(topfiveinjuries$EVTYPE))]
```

#### Top 5 Weather Events for Fatalities

Determine the top five weather event types for fatalities with a sum of all 
fatalities, grouped by event type, for the entire time period.

```{r, echo = TRUE, cache=TRUE}
fatalities <- health[,.(Fatalities=sum(FATALITIES)), EVTYPE]
setkey(fatalities, Fatalities)
topfivefatalities <- tail(fatalities, 5)
topfivefatalities[order(-Fatalities)]
```

From the list of top five weather event types for fatalities, calculate fatality 
sums by event type and month.

```{r, echo = TRUE, cache=TRUE}
fatalitiesbymonth <- health[EVTYPE %in% topfivefatalities$EVTYPE,
                        .(Injuries=sum(FATALITIES)), .(EVTYPE, Month)]
fatalitiesbymonth[, "EventType" := factor(EVTYPE, levels=rev(topfivefatalities$EVTYPE))]
```

### Economic Effects 

Filter property and crop damage for 50 US states and group by event type.

```{r, echo = TRUE, cache=TRUE}
library(datasets)
economic <- stormdata[STATE %in% state.abb, 
                      .(PROPDMG, CROPDMG, Month), by=EVTYPE]
```

#### Top 5 Weather Event Types for Damage

Determine the top five weather event types for property and crop damage with 
a sum of all damage, grouped by event type, for the entire time period.

```{r, echo = TRUE, cache=TRUE}
damage <- economic[,.(Damage=sum(
    sum(PROPDMG, na.rm=TRUE), sum(CROPDMG, na.rm=TRUE), na.rm=TRUE)/1e6), EVTYPE]
setkey(damage, Damage)
topfivedamage <- tail(damage, 5)
topfivedamage[order(-Damage)]
```

From the list of top five weather event types for crop and property damage, 
calculate damage sums by event type and month. Convert damage sums to 
millions of US dollars. 

```{r, echo = TRUE, cache=TRUE}
damagebymonth <- economic[EVTYPE %in% topfivedamage$EVTYPE,
    .(Damage=sum(
        sum(PROPDMG, na.rm=TRUE), sum(CROPDMG, na.rm=TRUE), na.rm=TRUE)/1e6), 
    .(EVTYPE, Month)]
damagebymonth[, "EventType" := factor(EVTYPE, levels=rev(topfivedamage$EVTYPE))]
```

## Results

### Public Health Effects

This plot shows the top five weather event types for injuries.

```{r, echo = TRUE, fig.height=5, fig.width=7}
library(ggplot2)
ggplot(injuriesbymonth, 
       aes(Month, Injuries, group=EventType, color=EventType)) +
    geom_line() +
    ggtitle("Top 5 Most Harmful\nWeather Event Types\n(Total US Injuries per Month, 1950-2011)") +
    ylab("Number of Injuries") +
    theme(plot.title = element_text(lineheight=.8, face="bold")) +
    scale_colour_brewer(palette="Set1") + 
    theme(legend.title=element_blank())
```

This plot shows the top five weather event types for fatalities.

```{r, echo = TRUE, fig.height=5, fig.width=7}
ggplot(fatalitiesbymonth, 
       aes(Month, Injuries, group=EventType, color=EventType)) +
    geom_line() +
    ggtitle("Top 5 Most Harmful\nWeather Event Types\n(Total US Fatalities per Month, 1950-2011)") +
    ylab("Number of Fatalities") +
    theme(plot.title = element_text(lineheight=.8, face="bold")) +
    scale_colour_brewer(palette="Set1") +
    theme(legend.title=element_blank())
```

#### Summary of Health Effects

Across the United States, springtime tornadoes are the most harmful to 
health, though heat, thunderstorms and lightning in midsummer also have a 
large impact. Tornadoes and flooding become the primary health hazards in 
the fall.

### Economic Effects

This plot shows the top five weather event types for economic impact.

```{r, echo = TRUE, fig.height=5, fig.width=7}
ggplot(damagebymonth, 
       aes(Month, Damage, group=EventType, color=EventType)) +
    geom_line() +
    ggtitle("Top 5 Most Harmful\nWeather Event Types\n(Total US Damage per Month, 1950-2011)") +
    ylab("Amount of Damage in Millions of US$") +
    theme(plot.title = element_text(lineheight=.8, face="bold")) +
    scale_colour_brewer(palette="Set1") + 
    theme(legend.title=element_blank())
```

#### Summary of Economic Effects

Across the United States, floods, especially in the winter and spring, 
cause the most economic damage, followed by hurricanes, typhoons, 
tornadoes, and storm surge. Hail also causes considerable damage, 
mostly in the spring.